---
title: "Dimensional reduction of phenotypes from 53,000 mouse models reveals a
        diverse landscape of gene function - Supplementary Figures"
output:
  pdf_document:
    keep_tex: true
header-includes:
  \usepackage{caption}
  \usepackage{float}
  \usepackage{amsmath}
  \usepackage{graphicx}
  \floatplacement{figure}{H}
  \captionsetup[figure]{labelfont={bf}, name={Figure}, labelsep=period}
---


```{r, analysis, echo=FALSE}
source("config.R")
set.seed(1234)
source("ME_prep_ontology.R")
source("ME_prep_models.R")
source("ME_prep_diseases.R")
source("ME_embeddings_ontology.R")
source("ME_embeddings_models.R")
source("ME_node2vec.R")
source("ME_predictions.R")
source("ME_genes.R")
set.seed(1234)
source("ME_embeddings_comparison.R")
```

```{r, graphics.details, echo=FALSE}
source("ME_wrappers.R")
# optionally set drawing to png
knitr::opts_chunk$set(dpi=600, dev="png", dev.args=list(type="cairo-png"))
#
# set detail for embedding plot
# set < 1 for quick rendering and =1 for final figures (png)
#EMBEDDING_DETAIL <- 0.1
EMBEDDING_DETAIL <- 1
```


\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}


```{r schematic_methods1, echo=FALSE}
plot_schematic_methods1 <- function() {
  plot_schematic_blank(height=0.8)
  # column with raw data
  add_schematic_line_header("Raw data", c(0, 0.3), 1)
  text(0, 0.92, "Ontology", Rcssclass=c("schematic", "subheading"))
  add_schematic_ontology(c(0.15, 0.87), width=0.16, height=0.14)
  text(0.15, 0.64, "+ term names, descriptions", Rcssclass=c("schematic", "center"))
  #
  # column with encoding cartoons
  add_schematic_line_header("Encoding", c(0.35, 0.65), 1)
  text(0.35, 0.92, "Text", Rcssclass=c("schematic", "subheading"))
  add_schematic_term_description(c(0.35,0.86),
                                 mp_raw[c("MP:0001967", "MP:0001393")],
                                 max.width=0.23, max.lines=5, indent=0.09)
  text(0.35, 0.4, "Graph", Rcssclass=c("schematic", "subheading"))
  add_schematic_ontology(c(0.5, 0.38), width=0.16, height=0.14)
  #
  # labels with embedding algorithms
  add_schematic_line_header("Embedding", c(0.7, 1), 1)
  text_algorithm <- function(...) { text(..., Rcssclass=c("schematic", "algorithm")) }
  text_algorithm(0.85, 0.7, "bag of kmers + nearest neighbors + UMAP")
  text_algorithm(0.85, 0.32, "node2vec")
  #
  # dividers
  lines(rep(0.325, 2), c(-1.5, 1), Rcssclass=c("schematic", "divider"))
  lines(c(0.35, 1), rep(0.47, 2), Rcssclass=c("schematic", "divider"))
}
```


```{r schematic_methods2, echo=FALSE}
plot_schematic_methods2 <- function() {
  plot_schematic_blank(height=1.35)
  # column with raw data
  add_schematic_line_header("Raw data", c(0, 0.3), 1)
  text(0, 0.92, "Ontology", Rcssclass=c("schematic", "subheading"))
  add_schematic_ontology(c(0.15, 0.87), width=0.16, height=0.14)
  text(0.15, 0.64, "+ term names, descriptions", Rcssclass=c("schematic", "center"))
  text(0, 0.49, "Model phenotypes", Rcssclass=c("schematic", "subheading"))
  dummy_phenotypes <- matrix("", ncol=2, nrow=4)
  colnames(dummy_phenotypes) <- c("model", "phenotype")
  dummy_phenotypes[,1] <- c("model A", "model B", "model B", "...")
  dummy_phenotypes[,2] <- c("MP:0001967", "MP:0001967", "MP:0001393", "...")
  add_schematic_table(c(0.15, 0.43), dummy_phenotypes, max.width=0.18)
  #
  # column with encoding cartoons
  add_schematic_line_header("Encoding", c(0.35, 0.65), 1)
  dummy_nonbinary <- matrix(0.1, ncol=10, nrow=2)
  rownames(dummy_nonbinary) <- c("model A  ", "model B  ")
  colnames(dummy_nonbinary) <- c("root", paste0("mid_", 1:3), paste0("leaf_", 1:6))
  dummy_binary <- dummy_nonbinary
  dummy_nonbinary[, "root"] <- 1
  dummy_nonbinary[, c("mid_1", "mid_2", "mid_3")] <- 0.4
  dummy_nonbinary["model A  ", "mid_1"] <- 0.7
  dummy_nonbinary["model B  ", "mid_1"] <- 0.7
  dummy_nonbinary["model B  ", "mid_2"] <- 0.6
  dummy_nonbinary["model A  ", "leaf_1"] <- 0.55
  dummy_nonbinary["model B  ", "leaf_1"] <- 0.55
  dummy_nonbinary["model B  ", "leaf_4"] <- 0.4
  dummy_binary[dummy_binary>0] <- 0
  dummy_binary["model A  ", c("root", "mid_1", "leaf_1")] <- 1
  dummy_binary["model B  ", c("root", "mid_1", "leaf_1", "mid_2", "leaf_4")] <- 1
  text(0.35, 0.92, "Vector (non-binary)", Rcssclass=c("schematic", "subheading"))
  add_schematic_heatmap(c(0.55, 0.87), dummy_nonbinary, width=0.16, height=0.09)
  text(0.35, 0.67, "Vector (binary)", Rcssclass=c("schematic", "subheading"))
  add_schematic_heatmap(c(0.55, 0.62), dummy_binary, width=0.16, height=0.09)
  #
  text(0.35, 0.42, "Graph", Rcssclass=c("schematic", "subheading"))
  model_A <- c(0.43, 0.33)
  text(model_A[1], model_A[2], "model A", Rcssclass=c("schematic", "model"))
  lines(c(model_A[1], 0.475), c(model_A[2], 0.27), Rcssclass=c("schematic", "ontology"))
  points(model_A[1], model_A[2], Rcssclass=c("schematic", "model"))
  model_B <- c(0.48, 0.42)
  text(model_B[1], model_B[2], "model B", Rcssclass=c("schematic", "model"))
  lines(c(model_B[1], 0.47), c(model_B[2], 0.27), Rcssclass=c("schematic", "ontology"))
  lines(c(model_B[1], 0.535), c(model_B[2], 0.27), Rcssclass=c("schematic", "ontology"))
  points(model_B[1], model_B[2], Rcssclass=c("schematic", "model"))
  add_schematic_ontology(c(0.55, 0.42), width=0.16, height=0.14)
  text(0.35, 0.17, "Text (concise phenotypes)", Rcssclass=c("schematic", "subheading"))
  dummy_concise <- list(A=list(data=list(name="deafness"),
                               metadata=list(id="model A")),
                        B=list(data=list(name="deafness, ataxia"),
                               metadata=list(id="model B")))
  add_schematic_term_description(c(0.4, 0.10), dummy_concise,
                                 max.width=0.23, max.lines=5, indent=0.06)
  text(0.35, -0.08, "Text (complete phenotypes)", Rcssclass=c("schematic", "subheading"))
  dummy_complete <- list(A=list(data=list(name="deafness",
                                          parents=c("abnormal ear physiology, ...")),
                                metadata=list(id="model A")),
                         B=list(data=list(name=c("deafness, ataxia"),
                                          parents=c("abnormal ear physiology, ...",
                                                    "abnormal locomotor behavior, ...")),
                                metadata=list(id="model B")))
  add_schematic_term_description(c(0.4, -0.14), dummy_complete,
                                 max.width=0.23, max.lines=5, indent=0.06)
  #
  # labels with embedding algorithms
  add_schematic_line_header("Embedding", c(0.7, 1), 1)
  text_algorithm <- function(...) { text(..., Rcssclass=c("schematic", "algorithm")) }
  text_algorithm(0.85, 0.86, "nearest neighbors + UMAP")
  text_algorithm(0.85, 0.79, "PCA")
  text_algorithm(0.85, 0.6, "nearest neighbors + UMAP")
  text_algorithm(0.85, 0.35, "node2vec")
  text_algorithm(0.85, 0.1, "bag of kmers + nearest neighbors + UMAP")
  text_algorithm(0.85, -0.2, "bag of kmers + nearest neighbors + UMAP")
  #
  # dividers
  lines(rep(0.325, 2), c(-1.5, 1), Rcssclass=c("schematic", "divider"))
  lines(c(0.35, 1), rep(0.73, 2), Rcssclass=c("schematic", "divider"))
  lines(c(0.35, 1), rep(0.48, 2), Rcssclass=c("schematic", "divider"))
  lines(c(0.35, 1), rep(0.23, 2), Rcssclass=c("schematic", "divider"))
  lines(c(0.35, 1), rep(-0.02, 2), Rcssclass=c("schematic", "divider"))
}
```


```{r supfig.mp, eval=TRUE, echo=FALSE, fig.width=6.5, fig.height=5.7, fig.cap="\\textbf{Embeddings of mammalian phenotype ontology terms.}"}
layout(matrix(c(1,1,1,1,1,1,1,
                2,2,2,2,2,2,2,
                3,7,4,8,9,9,9,
                5,7,6,8,9,9,9),
              ncol=7, nrow=4, byrow=T),
       widths=c(1.6, 0.04, 1.6, 0.04, 1.6, 0.04, 1.6),
       height=c(1.8, 0.7, 1.6, 1.6))
#
plot_schematic_methods1()
multipanelLabel(panel.labels[1])
#
wrap_separate_legend("bottomleft", c(other="MP term"), width=0.34,
                     main="Embeddings of mouse phenotype ontology terms")
#
# crossmap and umap embedding
plot_embedding(mp_embedding$d2_R0[label=="mp"],
               main=" Text descriptions",
               detail=max(0.5, EMBEDDING_DETAIL),
               Rcssclass="mp")
multipanelLabel(panel.labels[2])
#
# node2vec embedding
.n2v_encodings <- c("snap", "python_defaults", "python_short")
.encoding_main <- c(
  text="Text descriptions",
  snap="Graph (snap, default)",
  python_defaults="Graph (python, default)",
  python_short="Graph (python, short walks)"
)
for (.encoding in .n2v_encodings) {
  .main <- .encoding_main[.encoding]
  plot_embedding(node2vec_embedding$mp_ontology[[paste0(.encoding, "_R0")]][label=="mp",],
                 xy=c("node2vec_1", "node2vec_2"),
                 xlab="node2vec 1", ylab="node2vec 2",
                 main=paste0(" ", .main),
                 detail=max(0.5, EMBEDDING_DETAIL),
                 Rcssclass="mp")
  multipanelLabel(panel.labels[2+which(.n2v_encodings==.encoding)])
}
par(mai=rep(0.01, 4)); plot(c(0,1), c(0, 1), type="n")
par(mai=rep(0.01, 4)); plot(c(0,1), c(0, 1), type="n")
wrap_embedding_comparison(mp_embedding_comparison,
                          xlim=c(-2, 14), ylim=c(-14, 2),
                          labels=c(.encoding_main, "random"="Random layout"),
                          col.limits=c(0, 0.7),
                          legend.pos=c(0.5, -12, 8.5, -12.8))
rm(.encoding, .main, .n2v_encodings, .encoding_main)
multipanelLabel(panel.labels[6], x=-9, y=9)
```

\textbf{Embeddings of mammalian phenotype ontology terms.} \textbf{(A)} Schematic of ontology data, possibilities for numerical encoding, and algorithms for creating embeddings. \textbf{(B)} Embedding based on text descriptions (name, definition, synonyms, comments, and name of parent term). Similarities computed using crossmap and layout generated with UMAP. \textbf{(C)} Embedding based on the hierarchy relations between ontology terms, generated using snap implementation of node2vec with default settings. \textbf{(D)} Embedding based on the ontology hierarchy graph, generated using python implementation of node2vec with default settings. \textbf{(E)} Similar to E, generated with non-default settings with walk-length and num-walks set to 5.

\clearpage


```{r supfig.models.encodings, eval=TRUE, echo=FALSE, fig.width=6.5, fig.height=8.4, fig.cap="\\textbf{Embeddings of mouse models.}"}
layout(matrix(c(1,1,1,1,1,1,1,
                2,2,2,2,2,2,2,
                3,11,4,12,5,13,6,
                7,11,8,12, 14,14,14,
                9,11,10,12, 14,14,14),
              ncol=7, nrow=5, byrow=T),
       widths=c(1.6, 0.04, 1.6, 0.04, 1.6, 0.04, 1.6),
       height=c(2.9, 0.7, 1.6, 1.6, 1.6))
#
plot_schematic_methods2()
multipanelLabel(panel.labels[1])
#
wrap_separate_legend("bottomleft", c(other="mouse model"),
                     main="Embeddings of mouse models", width=0.22)
#
# sub-panels with different encodings
.encoding_main <- c(
  vector="Vector (non-binary)",
  pca="PCA (non-binary)",
  binvector="Vector (binary)",
  snap="Graph (snap)",
  python_defaults="Graph (python, default)",
  python_short="Graph (python, short walks)",
  text_concise_diff0="Text (concise)",
  text_complete_diff0="Text (complete)"
)
for (.encoding in names(.encoding_main)) {
  .main <- paste0(" ", .encoding_main[.encoding])
  if (paste0(.encoding, "_R0") %in% names(model_umap_embedding)) {
    plot_embedding(model_umap_embedding[[paste0(.encoding, "_R0")]][label=="model" & id %in% model_ids$phen_gt_1],
                   main=.main, detail=EMBEDDING_DETAIL,
                   Rcssclass=c("large", "models"))
  } else if (grepl("pca", .encoding)) {
    plot_embedding(model_pca_embeddings_d$vector$d2[label=="model" & id %in% model_ids$phen_gt_1],
                   xy=c("PCA_1", "PCA_2"),
                   xlab="PCA 1", ylab="PCA 2",
                   main=.main, detail=EMBEDDING_DETAIL,
                   Rcssclass=c("large", "models"))
  } else {
    plot_embedding(node2vec_embedding$mouse_model_concise[[paste0(.encoding, "_R0")]][label=="model" & id %in% model_ids$phen_gt_1],
                   xy=c("node2vec_1", "node2vec_2"),
                   xlab="node2vec 1", ylab="node2vec 2",
                   main=.main, detail=EMBEDDING_DETAIL,
                   Rcssclass=c("large", "models"))
  }
  multipanelLabel(panel.labels[1+which(names(.encoding_main)==.encoding)])
}
# filler panels
par(mai=rep(0.01, 4)); plot(1:10, type="n")
par(mai=rep(0.01, 4)); plot(1:10, type="n")
par(mai=rep(0.01, 4)); plot(1:10, type="n")
#
# embedding comparison panel
wrap_embedding_comparison(model_embedding_comparison,
                          labels=c(.encoding_main, "random"="Random layout"),
                          col.limits=c(0, 0.7),
                          legend.pos=c(0.5, -18, 8.5, -18.8))
multipanelLabel(panel.labels[10], x=-8, y=8)
rm(.encoding, .main, .encoding_main)
```

\textbf{Embeddings of mouse models.} \textbf{(A)} Schematic of phenotype data for mouse models, data encoding possibilities, and algorithms for creating embeddings. \textbf{(B)} UMAP embedding based on non-binary vector representations. \textbf{(C)} PCA based on non-binary vector representations. \textbf{(D)} UMAP embedding based on binary vector representations. \textbf{(E)} Node2vec embedding based on a graph connecting mouse models to their ontology phenotypes, generated using snap implementation of node2vec with default settings. \textbf{(F)} Similar to (E), generated using python implementation of node2vec with default settings. \textbf{(G)} Similar to (E), generated with python implementation of node2vec with non-default settings using walk-length and num-walks set to 5. \textbf{(H)} Embedding based on text descriptions of mouse phenotypes. \textbf{(I)} Similar to (H), but using text descriptions of complete phenotypes.

\clearpage


```{r supfig.models.covariates, eval=TRUE, echo=FALSE, fig.width=6.5, fig.height=6.5, fig.cap="\\textbf{Mouse model covariates.}"}
layout(matrix(c(1,3,10,
                2,4,10,
                5,5,10,
                6,8,11,
                7,9,11),
              ncol=3, nrow=5, byrow=T),
       widths=c(2.5, 2.5, 1.7),
       height=c(0.3+(0.09*2), 2.5, 0.1, 0.3+(0.09*7), 2.5))
#
# sub-panels with different encodings
.emb <- model_umap_embedding$vector_R0[label=="model"]
.emb <- merge(.emb, model_info, by.x="id", by.y="model_id")
discretize_base2 <- function(x) {
  result <- rep("range_0", length(x))
  result[x==1] <- "range_1"
  result[x==2] <- "range_2"
  result[x>2 & x<=4] <- "range_2_4"
  result[x>4 & x<=8] <- "range_4_8"
  result[x>8 & x<=16] <- "range_8_16"
  result[x>16] <- "range_16_max"
  result
}
discretize_bg <- function(x) {
  result <- rep("other", length(x))
  result[grep("Not.Specified", x)] <- "unspecified"
  result[grep("C57BL.6", x)] <- "C57BL_6"
  result[grep("C57BL.6N$", x)] <- "C57BL_6N"
  result[grep("C57BL.6NJ$", x)] <- "C57BL_6NJ"
  result[grep("C57BL.6NCrl$", x)] <- "C57BL_6NCrl"
  result[grep("C57BL.6NTac$", x)] <- "C57BL_6NTac"
  result
}
discretize_zygosity <- function(x) {
  result <- sapply(strsplit(x, "/"), function(z) {
    if (length(z)!=2) return("other")
    if (z[1]==z[2]) return("hom")
    return("other")
  })
  as.character(result)
}
.emb$discrete_phens <- discretize_base2(.emb$num_phenotypes)
.emb$discrete_bg <- discretize_bg(.emb$genetic_background)
.emb$discrete_zygosity <- discretize_zygosity(.emb$model_description)
#
# data sources (MGI/IMPC)
plot_legend_bars(counts=.emb[, list(count=nrow(.SD)), by="bin_source"],
                 labels=c(MGI="MGI mouse model", nonMGI="IMPC mouse model"),
                 xlab="Number of models", main="Data source")
multipanelLabel(panel.labels[1])
plot_embedding(.emb, style_by="bin_source", main="",
               detail=EMBEDDING_DETAIL, Rcssclass=c("large", "vlarge", "notop"))
#
# number of phenotypes
plot_legend_bars(counts=.emb[, list(count=nrow(.SD)), by="discrete_zygosity"],
                 labels=c(hom="homozygous", other="other"),
                 xlab="Number of models", main="Zygosity")
multipanelLabel(panel.labels[2])
plot_embedding(.emb, style_by="discrete_zygosity", main="",
               detail=EMBEDDING_DETAIL, Rcssclass=c("large", "vlarge", "notop"))
#
par(mai=rep(0.01, 4)); plot.new()
#
#  genetic background
plot_legend_bars(counts=.emb[, list(count=nrow(.SD)), by="discrete_bg"],
                 labels=c(C57BL_6="involves C57BL/6", C57BL_6N="C57BL/6N",
                          C57BL_6NCrl="C57BL/6NCrl", C57BL_6NJ="C57BL/6NJ",
                          C57BL_6NTac="C57BL/6NTac",
                          other="other", unspecified="not specified"),
                 xlab="Number of models", main="Genetic background",
                 Rcssclass="long")
multipanelLabel(panel.labels[3])
plot_embedding(.emb, style_by="discrete_bg", main="",
               style_groups=c("C57BL_6", "other", "unspecified",
                              "C57BL_6N", "C57BL_6BCrl", "C57BL_6NJ", "C57BL_6NTac"),
               detail=EMBEDDING_DETAIL, Rcssclass=c("large", "vlarge", "notop"))
#
# zygosity
plot_legend_bars(counts=.emb[, list(count=nrow(.SD)), by="discrete_phens"],
                 labels=c(range_0="0", range_1="1", range_2="2",
                          range_2_4="(2, 4]",
                          range_4_8="(4, 8]", range_8_16="(8, 16]",
                          range_16_max="(16, inf]"),
                 xlab="Number of models", main="Num. phenotypes",
                 Rcssclass="long")
multipanelLabel(panel.labels[4])
plot_embedding(.emb, style_by="discrete_phens", main="",
               detail=EMBEDDING_DETAIL, Rcssclass=c("large", "vlarge", "notop"))
#
#
par(mai=rep(0.01, 4)); plot.new()
#
#
.model_phen1 <- model_info[num_phenotypes==1]$model_id
.phen1_counts <- table(model_phenotypes$concise[id %in% .model_phen1]$phenotype)
.phen1_counts <- data.table(id=names(.phen1_counts),
                            count=as.integer(.phen1_counts))
.phen1_counts <- merge(.phen1_counts, mp_info$names, by="id")[order(-count)]
.phen1_counts$text <- paste0(.phen1_counts$count, " - ", .phen1_counts$name)
plot_list(.phen1_counts$text, xlim=c(0, 1), show.rect=FALSE,
          Rcssclass="counts")
text(0, 1.06, "Single phenotypes", adj=c(0, 0.5),
     Rcssclass=c("list", "main"))
multipanelLabel(panel.labels[5])
rm(.emb, .model_phen1, .phen1_counts)
```

\textbf{Mouse model covariates.} Embeddings of mouse models based on non-binary vectors of phenotypes. Four panels differ by stratification strategy: \textbf{(A)} phenotyping source; \textbf{(B)} zygosity of gene knock-out; \textbf{(C)} animal genetic background; \textbf{(D)} number of phenotypes. \textbf{(E)} Listing of the most frequent MP terms present among models with a single phenotype. Leading numbers indicate the number of models that are annotated with the stated term and no others.

\clearpage


```{r supfig.models.enrichment, eval=TRUE, echo=FALSE, fig.width=6.5, fig.height=7.8, fig.cap="\\textbf{Feature enrichment in embedding regions.}"}
layout(matrix(c(1,1, 16, 2, 2,
                3,3, 16, 3, 3,
                4,7, 16, 10,13,
                5,8, 16, 11,14,
                6,9, 16, 12,15),
              ncol=5, nrow=5, byrow=T),
       widths=c(1.6, 1.6, 0.1, 1.6, 1.6), height=c(3.2, 0.1, 1.5, 1.5, 1.5))
#
# embedding based on vectors
.emb1 <- model_umap_embedding$vector_R0[label=="model" & id %in% model_ids$phen_gt_1]
.emb1 <- merge(.emb1, model_info[, c("model_id", "bin_source")],
              by.x="id", by.y="model_id")
.emb1_gate1 <- c(-16, 5.5, -13, 8.5)
.emb1_gate2 <- c(-10, -1, -7, 2)
.qlim <- c(0.015, 0.985)
#
plot_embedding(.emb1, show_points=FALSE, qlim=.qlim,
               main=" Embedding of mouse models\n (non-binary vectors of MP terms)",
               Rcssclass=c("large", "vlarge", "title2"))
points_embedding(.emb1, detail=EMBEDDING_DETAIL, style_by="bin_source",
                 Rcssclass="large")
rect(.emb1_gate1[1], .emb1_gate1[2], .emb1_gate1[3], .emb1_gate1[4],
     Rcssclass="embedding")
rect(.emb1_gate2[1], .emb1_gate2[2], .emb1_gate2[3], .emb1_gate2[4],
     Rcssclass="embedding")
add_embedding_legend("bottomleft",
                     labels=c(MGI="MGI mouse model", nonMGI="IMPC mouse model"))
multipanelLabel(panel.labels[1])
#
# another embedding based on text
.emb2 <- model_umap_embedding$text_concise_diff0_R0[label=="model" &
                                                        id %in% model_ids$phen_gt_1]
.emb2 <- merge(.emb2, model_info[, c("model_id", "bin_source")],
              by.x="id", by.y="model_id")
.emb2_gate1 <- c(-5.5, -4, -2.5, -1)
.emb2_gate2 <- c(7, 2.5, 10, 5.5)
#
plot_embedding(.emb2, show_points=FALSE, qlim=.qlim,
               main=" Embedding of mouse models\n (text of concise phentypes)",
               Rcssclass=c("large", "vlarge", "title2"))
points_embedding(.emb2, detail=EMBEDDING_DETAIL, style_by="bin_source",
                 Rcssclass=c("large", "vlarge"))
rect(.emb2_gate1[1], .emb2_gate1[2], .emb2_gate1[3], .emb2_gate1[4],
     Rcssclass="embedding")
rect(.emb2_gate2[1], .emb2_gate2[2], .emb2_gate2[3], .emb2_gate2[4],
     Rcssclass="embedding")
add_embedding_legend("bottomleft",
                     labels=c(MGI="MGI mouse model", nonMGI="IMPC mouse model"))
multipanelLabel(panel.labels[6])
#
# spacer
par(mai=rep(0.01, 4)); plot.new();
#
# gate 1
wrap_embedding_enrichment(.emb1, gate_rect=.emb1_gate1,
                          panel.label=panel.labels[2:3],
                          xlim=c(0, 1), ylim=c(0, 320),
                          show_list=TRUE)
# gate 2
wrap_embedding_enrichment(.emb1, gate_rect=.emb1_gate2,
                          panel.label=panel.labels[4:5],
                          xlim=c(0, 1), ylim=c(0, 320),
                          show_list=TRUE)
# gate 3
wrap_embedding_enrichment(.emb2, gate_rect=.emb2_gate1,
                          panel.label=panel.labels[7:8],
                          xlim=c(0, 1), ylim=c(0, 320),
                          show_list=TRUE)
# gate 4
wrap_embedding_enrichment(.emb2, gate_rect=.emb2_gate2,
                          panel.label=panel.labels[9:10],
                          xlim=c(0, 1), ylim=c(0, 320),
                          show_list=TRUE)
rm(.emb1, .emb2, .emb1_gate1, .emb1_gate2, .emb2_gate1, .emb2_gate2, .qlim)
```

\textbf{Feature enrichment in embedding regions.} \textbf{(A)} Embedding of all mouse models based on their phenotypes. Two regions are selected with rectangles. \textbf{(B)} A detailed view of one of the selected regions from (A). \textbf{(C)} Enrichment analysis comparing phenotypes observed in animal models shown in (B) compared to all other models outside the selected region. Dots correspond to MP phenotypes. Axes show the fraction of selected models with a given phenotype, and an enrichment significance score for the phenotype computed using a Fisher test (some p-values are truncated). Significance level is $p=0.05$ after Bonferroni correction. The table on the right names the most significant phenotypes. \textbf{(D, E)} Analogous to (B, C), but treating another region. \textbf{(F-J)} Analogous to (A-J), but based on an embedding of mouse models based on concise text descriptions.

\clearpage


```{r supfig.genes, eval=TRUE, echo=FALSE, fig.align="center", fig.width=6.5, fig.height=6.5, fig.cap="\\textbf{Effect of genetic background for selected genes.} All panels show embeddings of mouse models based on non-binary vectors. In each panel, models with a single gene are highlighted: \\textbf{(A)} Tyr, \\textbf{(B)} Trp53, \\textbf{(C)} Kit, and \\textbf{(D)} Ednrb. One set of highlighted models reveals the most common genetic background (for that gene). The other highlighted set includes models with other genetic backgrounds (for that gene). The position of highlighted models are jittered to better reveal the number of models in dense areas."}
layout(matrix(c(1,6,2,
                5,5,5,
                3,7,4),
              ncol=3, nrow=3, byrow=T),
       widths=c(3.2, 0.1, 3.2), height=c(3.2, 0.1, 3.2))
#
.emb <- model_umap_embedding$vector_R0[label=="model" & id %in% model_ids$phen_gt_0]
.emb <- merge(.emb, model_info[, c("model_id", "marker_id", "marker_symbol", "genetic_background")],
              by.x="id", by.y="model_id")
wrap_embedding_gene(.emb, marker_symbols=c("Tyr", "Trp53", "Kit", "Ednrb"),
                    detail=EMBEDDING_DETAIL,
                    main="",
                    noise_sd=0.35,
                    highlight_dominant_group="genetic_background",
                    legend_pos="topleft",
                    panel.labels=panel.labels[1:4],
                    Rcssclass=c("large", "vlarge"))
#
# filler panels
par(mai=rep(0.01, 4)); plot.new();
par(mai=rep(0.01, 4)); plot.new();
par(mai=rep(0.01, 4)); plot.new();
```


```{r supfig.models.diseases, eval=TRUE, echo=FALSE, fig.width=6.5, fig.height=8.3, fig.cap="\\textbf{Embeddings of diseases.}"}
.disease_layout <- matrix(c(1,1,1, # header
                            2,3,4, # legend
                            2,3,5,  # xlab
                            2,3,6, # source
                            2,3,7, # zygosity
                            2,3,8, # genetic background
                            10,10,8, #
                            10,10,9), # number of phenotypes
                           nrow=8, ncol=3, byrow=TRUE)
layout(rbind(.disease_layout,
             matrix(rep(11, 3), ncol=3),
             .disease_layout+11),
       widths=c(2.15, 2.15, 2.2),
       height=c(0.2, 0.5, 0.2, 0.46, 0.46, 0.73, 0.38, 1.11,
                0.22,
                0.2, 0.5, 0.2, 0.46, 0.46, 0.73, 0.38, 1.11))
#
# vector embedding, sub-panels with diseases
.emb1 <- model_umap_embedding$vector_R0
.emb1_models <- .emb1[label=="model"]
.disease_ids <- .emb1[label=="owlsim"]$id
.qlim <- c(0.013, 0.987)
#
make_discrete_summary <- function(ids, label="") {
  ids <- unique(ids[!is.na(ids)])
  result <- model_info[model_id %in% ids]
  result$discrete_phens <- discretize_base2(result$num_phenotypes)
  result$discrete_bg <- discretize_bg(result$genetic_background)
  result$discrete_zygosity <- discretize_zygosity(result$model_description)
  result <- result[, c("model_id", "discrete_phens", "discrete_zygosity",
                       "discrete_bg", "bin_source")]
  n_models <- nrow(result)
  result <- melt(result, id.vars="model_id")
  result <- result[, list(count=nrow(.SD), percent=100*nrow(.SD)/n_models),
                     by=c("variable", "value")]
  result$label <- label
  setcolorder(result, c("label", "variable", "value", "count"))
  result
}
.info1 <- rbind(
        make_discrete_summary(model_ids$all, label="all"),
        make_discrete_summary(model_vectors_ids[as.vector(disease_model_knn$owlsim$vector$indexes)],
                              label="owlsim"),
        make_discrete_summary(model_vectors_ids[as.vector(disease_model_knn$crossmap$vector$indexes)],
                              label="crossmap")
)
#
#
plot_line_header("Analyses based on non-binary vectors of MP terms");
#
plot_embedding(.emb1_models[id %in% model_ids$phen_gt_1],
               detail=max(0.1, EMBEDDING_DETAIL),
               main="", qlim=.qlim, Rcssclass=c("large", "large3"))
points_embedding(.emb1[label=="owlsim"], detail=max(0.2, EMBEDDING_DETAIL),
                 Rcssclass="owlsim")
add_embedding_legend(legend_pos="topleft", legend_shift=c(-1, 5.5),
                     labels=c(other="Mouse model",
                              owlsim="Human disease (owlsim HP-MP transl.)")
)
multipanelLabel(panel.labels[1])
plot_embedding(.emb1_models[id %in% model_ids$phen_gt_1],
               detail=max(0.1, EMBEDDING_DETAIL),
               main="", qlim=.qlim, Rcssclass=c("large", "large3"))
points_embedding(.emb1[label=="crossmap"], detail=max(0.2, EMBEDDING_DETAIL),
                 Rcssclass="crossmap")
add_embedding_legend(legend_pos="topleft", legend_shift=c(-1, 5.5),
                     labels=c(other="Mouse model",
                              crossmap="Human disease (crossmap HP-MP transl.)")
)
multipanelLabel(panel.labels[2])
#
par(mai=c(0.02, 0.2, 0.12, 0.2)); plot(c(0, 1), c(0, 1), type="n", axes=FALSE);
text(0, 1, "Properties of mouse models that appear\nas neighbors (k=15) to human diseases",
     Rcssclass=c("embedding", "label"), adj=c(0, 0.5))
add_embedding_legend(legend_pos=c(0.0, 0.8),
                     labels=c(owlsim="owlsim transl.", crossmap="crossmap transl."))
add_embedding_legend(legend_pos=c(0.5, 0.8), labels=c(all="All models"))
multipanelLabel(panel.labels[3])
wrap_disease_models(.info1)
#
par(mai=rep(0.1, 4)); plot(c(0, 1), c(0, 1), type="n", axes=FALSE);
#
#
# text-based embeddings
assignc("disease_search")
.crossmap_search <- disease_search$crossmap$orphanet$text_concise_diff0
.owlsim_search <- disease_search$owlsim$orphanet$text_concise_diff0
.disease_search_1 <- list(
  crossmap = .crossmap_search[, head(.SD, 1), by="query"],
  owlsim = .owlsim_search[, head(.SD, 1), by="query"]
)
.emb2 <- model_umap_embedding$text_concise_diff0_R0
.emb2_models <- .emb2[label=="model"]
.emb2_diseases <- lapply(
  .disease_search_1,
  function(x) {
    x$id <- x$query
    predict_avg_coordinates(.emb2, x, feature_col="target", label="disease")
  }
)
.info2 <- rbind(
        make_discrete_summary(model_ids$all, label="all"),
        make_discrete_summary(.crossmap_search[query %in% disease_ids$phen_eq_0]$target,
                              label="crossmap_phen_eq_0"),
        make_discrete_summary(.crossmap_search[query %in% disease_ids$phen_gt_0]$target,
                              label="crossmap_phen_gt_0"),
        make_discrete_summary(.owlsim_search[query %in% disease_ids$phen_gt_0]$target,
                              label="owlsim")
)
#
par(mai=rep(0.02, 4)); plot(c(0, 1), c(0, 1), type="n", axes=FALSE);
#
plot_line_header("Analyses based on text descriptions");
# row with owlsim results
plot_embedding(.emb2_models[id %in% model_ids$phen_gt_1],
               detail=max(0.1, EMBEDDING_DETAIL),
               main="", qlim=.qlim, Rcssclass=c("large", "large3"))
points_embedding(.emb2_diseases$owlsim[id %in% .disease_ids],
                 detail=max(0.2, EMBEDDING_DETAIL),
                 Rcssclass="owlsim")
add_embedding_legend(legend_pos="topleft", legend_shift=c(-0.5, 3.5),
                     labels=c(other="Mouse model",
                              owlsim="Human disease (owlsim HP-MP transl.)"))
multipanelLabel(panel.labels[4])
#
plot_embedding(.emb2_models[id %in% model_ids$phen_gt_1],
               detail=max(0.1, EMBEDDING_DETAIL),
               main="", qlim=.qlim, Rcssclass=c("large", "large3"))
points_embedding(.emb2_diseases$crossmap[id %in% .disease_ids],
                 detail=max(0.2, EMBEDDING_DETAIL),
                 Rcssclass="crossmap")
add_embedding_legend(legend_pos="topleft", legend_shift=c(-0.5, 3.5),
                     labels=c(other="Mouse model",
                              crossmap="Human disease (crossmap HP-MP transl.)"))
multipanelLabel(panel.labels[5])
#
par(mai=c(0.02, 0.2, 0.12, 0.2)); plot(c(0, 1), c(0, 1), type="n", axes=FALSE);
text(0, 1, "Properties of mouse models that appear\nas neighbors (k=15) to human diseases",
     Rcssclass=c("embedding", "label"), adj=c(0, 0.5))
add_embedding_legend(legend_pos=c(0.0, 0.8),
                     labels=c(owlsim="owlsim transl.",
                              crossmap_phen_gt_0="crossmap transl."))
add_embedding_legend(legend_pos=c(0.5, 0.8),
                     labels=c(crossmap_phen_eq_0="descriptions only",
                              all="All models"))
multipanelLabel(panel.labels[6])
wrap_disease_models(.info2)
#
#
par(mai=rep(0.1, 4)); plot(c(0, 1), c(0, 1), type="n", axes=FALSE);
#
rm(.emb1, .emb1_models, .disease_ids, .qlim)
rm(.emb2, .emb2_models, .emb2_diseases)
rm(.crossmap_search, .owlsim_search, .disease_search_1)
```

\textbf{Embeddings of diseases.} \textbf{(A)} Projection of human disease phenotype profiles (colored points) into an embedding of mouse models (gray dots). The embedding was created based on non-binary vectors with mouse model phenotypes, without information about diseases. Disease phenotype profiles were translated from the human phenotype (HP) into the mammalian phenotype (MP) ontology terms using owlsim, encoded into vectors, and projected into the embedding using UMAP. \textbf{(B)} Similar to (A), but with the translations between human and mammalian ontologies carried out using crossmap. \textbf{(C)} Comparison of the properties of mouse models identified during disease analysis to the properties of the entire mouse model cohort. Disease profiles were compared with all mouse models, recording 15 nearest mouse models for each disease. The set of thus selected models were stratified according to the mouse model data source, mutation zygosity, mouse strain/genetic background, and the number of phenotypes annotated to the mouse models. The same stratification was applied to the entire set of mouse models for comparison. \textbf{(D, E, F)} Analogous to panels (A, B, C), but with all analyses carried out using text-based similarities. Analysis in panel (F) includes a group of models identified during analysis of diseases that only have text descriptions, without usage of formal ontology-based phenotypes.

\clearpage


```{r supfig.hpmp, eval=TRUE, echo=FALSE, fig.align="center", fig.width=5.2, fig.height=5.3, out.width="80%", fig.cap="\\textbf{Disease phenotypes in embeddings of mammalian and human phenotype ontologies.} \\textbf{(A)} Embedding of terms from the mammalian phenotype (MP) ontology; dots are colored by whether or not MP terms appear in (translated) disease definitions. \\textbf{(B)} Embedding of MP terms from (A) with additional points representing disease profiles computed through MP coordinate averaging. \\textbf{(C)} Embedding of terms from the human phenotype (HP) ontology; dots are colored by whether or not HP terms appear in disease annotation (without translation). \\textbf{(D)} Embedding of HP terms from (C) with additional points representing disease profiles computed by HP coordinate averaging."}
layout(matrix(c(1,2,
                3,3,
                4,5),
              ncol=2, nrow=3, byrow=T),
       widths=c(2.6, 2.6), height=c(2.6, 0.1, 2.6))
#
.emb_mp <- mp_embedding$d2_R0
.emb_hp <- hp_embedding$d2_R0
#
# panel with individual phenotypes
plot_embedding(.emb_mp[label=="mp"],
               main="", qlim=c(0.005, 0.995),
               detail=1,
               Rcssclass="mp")
points_embedding(.emb_mp[id %in% disease_phenotypes$crossmap$phenotype],
                 detail=1, Rcssclass="crossmap")
add_embedding_legend("topleft",
                     labels=c(crossmap="MP term appearing in a disease", mp="other MP term"))
multipanelLabel(panel.labels[1])
#
# panel with disease profiles (by coordinate averaging)
plot_embedding(.emb_mp[label=="mp"],
               main="", qlim=c(0.005, 0.995),
               detail=1,
               Rcssclass="mp")
points_embedding(.emb_mp[label=="crossmap"],
                 detail=1, Rcssclass="crossmap")
add_embedding_legend("topleft",
                     labels=c(crossmap="disease profile", mp="MP term"))
multipanelLabel(panel.labels[2])
#
par(mai=rep(0.01, 4)); plot.new();
#
#
# panel with individual phenotypes
plot_embedding(.emb_hp[label=="hp"], detail=1,
               main="", qlim=c(0.005, 0.995),
               Rcssclass="mp")
points_embedding(.emb_hp[id %in% disease_phenotypes_hp$phenotype], detail=1,
                 Rcssclass="disease_hp")
add_embedding_legend("topleft",
                     labels=c(disease_hp="HP term appearing in a disease", mp="other HP term"))
multipanelLabel(panel.labels[3])
#
# panel with disease profiles (by coordinate averaging)
plot_embedding(.emb_hp[label=="hp"], detail=1,
               main="", qlim=c(0.005, 0.995),
               Rcssclass="mp")
points_embedding(.emb_hp[label=="disease"], detail=1,
                 Rcssclass="disease_hp")
add_embedding_legend("topleft",
                     labels=c(disease_hp="disease profile", mp="HP term"))
multipanelLabel(panel.labels[4])
#
rm(.emb_mp, .emb_hp)
```

